Temporal Evolution of Erosion in Pore-Networks:\\ From Homogenization to Instability

### 2022-03-06 review_response/Response.tex

%%//////////////////////////////////////////////////////////////
%%
%% P R E A M B L E
%%
%%//////////////////////////////////////////////////////////////
\documentclass{article}

%%==============================================================
%% Packages=
%%==============================================================
\usepackage[english]{babel}
\usepackage{booktabs}
\usepackage{scicite}
%\usepackage{citesort}
\usepackage[usenames,dvipsnames]{color}
\usepackage{graphics,amsfonts,amssymb,amsmath,latexsym}
\usepackage{graphicx,subfig}
\usepackage[ansinew]{inputenc}
%\usepackage{natbib}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{subfig}
\usepackage{textcomp}
\usepackage{hyperref}
\usepackage{lscape}
\usepackage{authblk}
\usepackage{xr}
\usepackage{enumerate}
\usepackage{bm}
\usepackage{epstopdf}
\usepackage{float}
\setlength\parindent{0pt}
\usepackage{xr}
\usepackage{csquotes}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor}

\usepackage[usenames, dvipsnames]{color}
\usepackage[dvipsnames]{xcolor}

% Color Edits
\newcommand{\add}[1]{\noindent \color{blue} #1 \normalcolor}
% \newcommand{\del}[1]{\noindent \color{red} \st{#1}\normalcolor}
% \newcommand{\AM}[1]{\noindent \color{magenta} (AM: #1)\normalcolor}
% \newcommand{\AZ}[1]{\noindent \color{purple} (AZ: #1)\normalcolor}
% \newcommand{\PD}[1]{\noindent \color{cyan} (PD: #1)\normalcolor}

%\usepackage{scicite}
\usepackage{natbib}
\bibliographystyle{unsrt}

%%==============================================================
%% New Commands
%%==============================================================
\newcommand{\Hline}{\rule{\linewidth}{.1mm}}
%\newcommand{\RedColor}[1]{\noindent \color{red}  #1 \normalcolor}
\newcommand{\Question}[1]{\noindent \color{black}\emph{#1}\normalcolor}
\newcommand{\Answer}[1]{\noindent {\color{blue}{ #1}}\normalcolor}

\newcommand{\AnswerQ}[1]{\noindent {\footnotesize {\color{blue}{ #1}}}\normalcolor}

\renewcommand{\thepage}{R\arabic{page}}
\renewcommand{\thesection}{R\arabic{section}}
\renewcommand{\theequation}{R\arabic{equation}}
\renewcommand{\thefigure}{R\arabic{figure}}
\renewcommand{\thetable}{R\arabic{table}}
\renewcommand{\thetable}{R\arabic{table}}

\def\th{\theta}

\newcommand{\bfphi}{\boldsymbol{\phi}}%

\newcommand{\Hb}{\mathbf{\bar{H}}}

\newcommand{\revised}[1]{\noindent\color{Blue}#1\normalcolor}

\newcommand{\PD}[1]{\noindent \color{red} (PD: #1)\normalcolor}
\newcommand{\EM}[1]{\noindent \color{magenta} (EM: #1)\normalcolor}
\newcommand{\AZ}[1]{\noindent \color{purple}(AZ: #1)\normalcolor}

\def\p{\textit{p.~}}
\def\l{\textit{l.~}}
%%==============================================================
%% Setttings
%%==============================================================
\topmargin -0.5in

\setlength{\textwidth}{6in} \setlength{\textheight}{8.25in}
\setlength{\oddsidemargin}{0mm}
\setlength{\textwidth}{6.in}
\setlength{\textheight}{8.5in}
\setlength{\evensidemargin}{0.25in}
\setlength{\oddsidemargin}{0.25in}
\setlength{\oddsidemargin}{0mm}

%%//////////////////////////////////////////////////////////////
%%
%% T I T L E    &    A U T H O R S
%%
%%//////////////////////////////////////////////////////////////
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{
Response to the Reviewers' comments for manuscript \\
(Tracking \#: LF18481): \\
``Temporal evolution of flow in pore-networks: From homogenization to instability"
}

\author{Ahmad Zareei, Deng Pan, and  Ariel Amir
}
\date{}

%%//////////////////////////////////////////////////////////////
%%
%% FIRST SENTENSE
%%
%%//////////////////////////////////////////////////////////////
\begin{document}
\maketitle

\noindent In the following, we address each of the Referees' comments/suggestions (\emph{in italic}). All page/paragraph/line (\textit{p.},\ \textit{par.},\ \textit{l.},\ etc.) numbers refer to the revised paper. %In addition, while using the original numbering of figures/equations/references in the manuscript (\emph{e.g.}, Fig.~1, Fig.~2,...., Eq.~(1), Eq.~(2),..., [1], [2], ...),  we add prefix ``\emph{R}'' for figures/equations/references presented in this response to the Referees' comments (\emph{e.g.}, Fig.~R1, Fig.~R2, .... Eq.~(R1), Eq.~(R2), ..., Ref.~[R1], Ref.~[R2],...).
% \centerline{(Tracking \#: LB15473 Deng)}

\vspace{5 mm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Ref 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent
\Hline \\
\textbf{Response to Reviewer \#1} \\
\Hline
\\

% \centerline{Negative numbers are counted from the bottom of the page.}
%({\color{orange}\mbox{------}})

\Question{
This paper studies the evolution of a porous medium network model
under erosion. The basic result, which does not require any
calculation at all, is that when the wider channels are more likely to erode the flow becomes localized and when erosion is stronger in the smaller channels the flow becomes uniform. This is because in pore networks most of the flux is in the wider channels so if these are more likely to widen they will carry more and more flux, whereas more erosion in the narrow channels allows them to carry more flux and evens out the distribution.
% 
The calculations are of theoretical interest but not significant
enough for PRL. The paper could be revised to combine the
supplementary material with the main text and considered for
publication in PRE or PRFluids.
\newline}

\Answer{{We thank the reviewer for carefully reading our manuscript and providing insightful comments.}}

\vspace{0.5cm}
\textbf{Comment 1A }
\noindent \vspace{-0.2cm}\\ \Hline\\

\Question{The particular model introduced in Eq. 1 gives a local erosion rate proportional to arbitrary powers of the pressure difference and the radius, which is hard to rationalize outside of the simple case $m=1$ and $n=0$. It is probably useful to have specific models illustrating an obvious result but they are too generic to be quantitatively applicable; a justification based on a serious analysis of the erosion process would be needed. \newline}

\Answer{We thank the reviewer for encouraging us to consider more realistic models of physical erosion. To address this issue, in the revised version, we have taken different steps to make our models more realistic: (i) We identify the stability condition for the homogeneous phase starting from a rather general equation for the erosion dynamics, of which the previous power-law class is a particular case. (ii) We show that for a more realistic model of erosion -- where the erosion rate depends on the shear stresses in a nonlinear fashion, effectively incorporating both a threshold for the onset of erosion as well as saturation of the erosion rate -- our approach correctly predicts the onset of channeling and the fate of the network (which in this case may, interestingly, result in multiple channels). (iii) We further show that the model considered in the manuscript is robust to the form of the boundary conditions, in addition to the robustness to topology and the form of the disorder that was previously considered. (iv) Following previous models proposed for erosion in the literature, we consider an extension of our previous model where the onset of erosion occurs only after reaching a threshold value, while still maintaining the previously used power-law form thereafter.  We analytically show that the phase transition persists with a slightly modified condition, which we corroborate numerically. (v) Finally, we expanded our discussion of erosion models and elaborate on the connection with the existing models in literature. These results are now included both in the main text and new sections of the supplementary material. We attach the relevant excerpts from the text below.}
%
\\

\Answer{$\bullet$ (i) From \p3 and \p4 of the main text:}
%
\AnswerQ{
\begin{quote}
    ``\textit{Local Dynamics Model}-- To understand the transition in network behavior during erosion, we focus on a simplified model with only two tubes in parallel or series with a general erosion dynamics as $dr_{i}/dt = f(q_{i},r_{i}), i=1,2$ (Figs.~3b-c). First, assuming two cylindrical tubes with radii $r_1,r_2$ in series with a given pressure difference of $\delta p = p_r- p_l$, the radius of each tube changes as $dr_i/dt = f(q,r_i),~i=1,2$ (Fig. 3b) where $q=q_1=q_2$. Without loss of generality we assume $r_1>r_2$. Considering the pressure at the junction between tubes $\tilde p_m = (p_m-p_l)/(p_r-p_l)= C_2/(C_1+C_2)$. Time evolution of the middle pressure then follows % 
    %
    \begin{align}
        \frac{d\tilde{p}_m}{dt} = \gamma (r_1 f(q,r_2)-r_2 f(q,r_1)). 
    \end{align}
    %
    where $\gamma = \sqrt{{2\pi}/{\mu L}}{C_1^{\frac{3}{4}} C_2^{\frac{3}{4}}}/{ (C_1+C_2)^2}$. In the above equation, $r_1=r_2$ is a fixed point solution where $d\tilde p_m/dt =0$ and $p_m=1/2$. Since we assumed $r_1>r_2$, we can see that $\tilde p_m <1/2$, and if ${d\tilde{p}_m}/{dt}>0$, then $\tilde p_m \to 1/2$ and the flow homogenizes. Note that if $f(q,r) = \alpha q^m/r^n$, we find that for any $n>0$ the network homogenizes (see Fig. 3b). Contrary to tubes in series, when the tubes are in parallel (Fig. 3c), the net flow $q$ divides between the two tubes in proportion to their conductivity, i.e., $ q_1/q_2 = C_1/C_2$ which results in $q_1/q = \tilde C_1$, where $\tilde C_1 = C_1/(C_1+C_2)$. Note that the dynamics for each pipe in the parallel case can be simplified to $dr_i/dt = g(r_i)$ where $g(r_i) \equiv f(q_i, r_i)$ in which $q_i = \pi r_i^4 \Delta p / 8\mu L$. Given these dynamics, the evolution of the fluid flow ratio becomes $d(q_1/q)/dt = d \tilde C_1/dt=  \gamma (r_2 g(r_1)-r_1 g(r_2))$. Similarly, it can be seen that  $r_1=r_2$ is a fixed point, where flow is equally distributed between the edges resulting in homogenization. If $d \tilde C_1/dt <0$, then the edge with larger radius has a reduced growth rate while the smaller radius edge has an increasing growth until $r_1=r_2$ is reached. This condition results in homogenization, where both edge's radius increases until the fixed point of $r_1=r_2$ is reached. The condition of $d \tilde C_1/dt<0$ is satisfied when 
    %
    \begin{align}
        \frac{g(r_1)}{g(r_2)}<\frac{r_1}{r_2}, \quad \forall r_1>r_2>0.   \label{Eq3}
    \end{align}
    %
    Note that if $d\left(g(r)/r\right)/dr<0$, then Eq.~\eqref{Eq3} is satisfied, and homogenization would occur.
    The above homogenization condition can further be simplified as $g(r)>g'(r)r$, which can also be obtained directly using a linear stability analysis. Given the homogenization condition $g(r)>g'(r)r$, if the erosion rate is sub-linear locally near the initial radius, the network moves towards homogenization. On the other hand, if the erosion rate is super-linear, we expect the parallel tubes to move away from homogenization and channelization occurs. Applying the power law erosion model with $f(q_{i},r_i)=\alpha q_i^m/r_i^n$, it can be shown that $g(r) \propto  r^{4m-n} $, and therefore $4m-n=1$ results in no change in the flow ratio $q_1/q$ (Fig. 3c$_1$), while $4m-n<1$ results in homogenization condition (Fig.~3c$_2$), and $4m-n>1$ results in channelization (Fig. 3c$_3$). Since a complex network includes both series and parallel connections, it is plausible that the whole network structure will behave in a similar manner, with an approximate transition in the network's behavior at $ m = (n+1)/4$. This result for $m=1$ reduces to a transition at $n=3$. It is to be noted that while the flow ratio remains constant for parallel edges, the connections in series are evolving toward homogenization and that is why we find a difference between evolved network for $n=3$ and the network at $t=0$. Since series connections are always moving toward homogenization, while parallel connections show a phase transition, we expect the phase transition to be at approximately $n\approx 3$, governed by the parallel connections. This observation  is consistent with the numerical simulation results shown in Figs.~2 and 3a (for $m=1$) as well as for additional values of $m$ (Fig.~4). Note, however, that the local dynamics model is only approximate and the numerically observed value of the phase transition happens for a value slightly larger than $n=3$ in Fig.~3.''
    \end{quote}
}
\\
\newline 
\vspace{4mm}

\Answer{$\bullet$ (ii) From supplementary material, section S7:}
%
\AnswerQ{
\begin{quote}
    ``As an example of a nonlinear erosion model, we consider a shear-dependent erosion
    %
    \begin{align}
        \frac{dr}{dt} = f(\tau), 
    \end{align}
    %
    where $\tau=\tau(q,r) = 4\mu q/\pi r^3$ is the shear stress at the tube's wall assuming Poiseuille flow. It is to be noted that a shear-dependent erosion model is commonly used in studying erosion  \cite{jager2017channelization,bonelli2011micromechanical,parker2000purely} wherein such models erosion happens linearly proportional to the shear stress at the wall only after a certain threshold value $\tau_0$ (see main text for more information). Such linear models, however, ignore the maximum detachment rate of particles at the pore's boundary which would effectively limit the pore's radius maximum rate of change. To address this, we use a nonlinear erosion model
    %
    \begin{align}
        f(\tau) = \alpha \left( \tanh{\left[ \beta(\tau - \tau_0)\right]}+1\right), 
        \label{eq:sigmoidErosion}
    \end{align}
    %
    where $\beta$ determines the shear rate scale and $\alpha$ determines the erosion rate scale. Note that the erosion rate scale $\alpha$ only affects the timescale at which the final state is reached and has no effect on the pattern formation. Without loss of generality we set $\alpha = 1$. Following the local dynamics model discussion in the main text, one can find that the homogenization condition reduces to 
    %
    \begin{align}
        f(\tau)>\tau \frac{df(\tau)}{d \tau}. 
        \label{eq:sigmoidCondition}
    \end{align}
    %
    Assuming a random network (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}a), the initial values of shear stress at the walls has a  decaying distribution over a large set of values (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}b). 
    % Assuming nonlinear shear dependent erosion model (Eq. \eqref{eq:sigmoidErosion}), and the homogenization condition (Eq. \eqref{eq:sigmoidCondition}), three regions of erosion can be identified: (i) constant erosion region as $\tau\to\infty$: at this region the erosion happens at a constant rate $\lim_{\tau \to \infty} f(\tau) = 2$ and additionally the homogenization condition (Eq.~\eqref{eq:sigmoidCondition}) is satisfied (since $\lim_{\tau \to \infty} f(\tau) = 2$ and $\lim_{\tau \to \infty} f'(\tau) = 0$). The homogenization results here is similar to the constant erosion rate model or $m=0,n=0$ discussed in the main text; (ii)  the slow erosion region as $\tau\to 0$. In this region both $f(\tau)$ and $\tau f'(\tau)$ are close to 0 and erosion happens at a very slow rate; (iii) the finite erosion-rate region near $\tau = \tau_0$, where $f(\tau)$ is an increasing function. In this region $\tau f'(\tau)$ has a maximum at $\beta \tau_0$ with a bandwidth of $1/\beta$. Note that here if $\beta \tau_0 < 1$, then the maximum $\tau f'(\tau)$ does not exceed 1, and the homogenization condition holds true for any $\tau$. As the value of $\tau_0$ increases, the homogenization region moves towards larger values of $\tau$ where smaller values of $\tau$ would result in channelization.  Varying $\beta$ and $\tau_0$ can change the network behavior. 
    To test our local dynamics model, we vary $\beta$ and $\tau_0$ and run three different tests: (i) $\beta=1,\tau_0=1$: For such values of $\beta,\tau$, almost all the tubes are in the homogenization region (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}c) and the network moves toward homogenization (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}d). The final PDF of the network after erosion is further shown in (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}e). (ii) $\beta=1, \tau_0=12$: In this case, the homogenization condition is satisfied at some finite value of shear stress $\tau_0$ where $f(\tau_0) = \tau_0 f(\tau_0)$. As a result, some edges would homogenize, and some would channelize (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}f). Running the simulation over a random network, we find that a finite number of channels form (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}g). The PDF distribution of shear stress at the wall $\tau$ correspondingly becomes bimodal (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}h). (iii) $\beta=1, \tau_0=30$: In this case, almost all the edges in the network are in the channelization regime  (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}i), and performing the simulations we find that indeed one strong single channel is formed (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}j) and the PDF of shear stress $\tau$ shows a clear separation between flow in the channel versus flow in the rest of the system (Fig.~\ref{SIfig:fig9-sigmoidDelaunay}k).''
\end{quote}
}

### 2022-03-06 CL/CoverLetter.tex
Updates from Overleaf

\signature{\vspace{-10mm}\includegraphics[height=0.4in]{Signature.eps} \\\textbf{Ariel Amir}\\ \textit{Thomas D. Cabot Associate Professor of Applied Mathematics and Applied Physics,} \\ \textit{School of Engineering and Applied Sciences,}\\ \textit{Harvard University} \\ \textit{Phone: (617) 495-5818}\\ \textit{Email: arielamir@seas.harvard.edu}}

@@

% \begin{table}
%     \centering
    \begin{tabular}{l l }
        \small{Title:}  & \textbf{\small{Temporal Evolution of flow in Porous Materials: From Homogenization to Instability}} \vspace{2mm} \\ 
        \small{Authors:}  & \small{Ahmad Zareei$^1$, Pan Deng$^1$, and Ariel Amir$^{1,2}$* (*corresponding author)} \vspace{2mm} \\
        \small{Affiliations:} & \small{$^1$Harvard John A. Paulson School of Engineering and Applied Sciences, Harvard University} \\
        & \small{$^2$Kavli Institute, Harvard University}
    \end{tabular}
% \end{table}

@@

We are excited to submit our work, ``Temporal Evolution of flow in Porous Materials: From Homogenization to Instability'' for consideration as an article in the \emph{Physical Review Letters}. In this study, we demonstrate that ....

As such, our study suggests that

@@

\textbf{Karen Alim},Max Planck Institute for Dynamics and Self-Organization, G\"ottingen, Germany, \texttt{k.alim@tum.de}

>>>>

\signature{% \vspace{-10mm}\includegraphics[height=0.4in]{Signature.eps}
\\\textbf{Ariel Amir}\\ \textit{Thomas D. Cabot Associate Professor of Applied Mathematics and Applied Physics,} \\ \textit{School of Engineering and Applied Sciences,}\\ \textit{Harvard University} \\ \textit{Phone: (617) 495-5818}\\ \textit{Email: arielamir@seas.harvard.edu}}

@@

% \begin{table}
%     \centering
    \begin{tabular}{l l }
        \small{Title:}  & \textbf{\small{Temporal Evolution of flow in Pore-Networks: From Homogenization to Instability}} \vspace{2mm} \\ 
        \small{Authors:}  & \small{Ahmad Zareei$^1$, Pan Deng$^1$, and Ariel Amir$^{1,2}$* (*corresponding author)} \vspace{2mm} \\
        \small{Affiliations:} & \small{$^1$Harvard John A. Paulson School of Engineering and Applied Sciences, Harvard University} \\
        & \small{$^2$Kavli Institute, Harvard University}
    \end{tabular}
% \end{table}

@@

We are excited to submit our work “Temporal Evolution of flow in Pore-Networks: From Homogenization to Instability" for consideration as an article in Physical Review Letters.
In this manuscript, we study the temporal evolution of flow in networks under dynamic changes in the pore structure, using a combination of analytical and numerical approaches.

Fluid flow through porous materials undergoing a dynamical change in its network of microstructure due to erosion or material deposition is ubiquitous in nature and has numerous environmental and industrial applications. Understanding the dynamical change is essential to improve any of the applications. In light of the plethora of applications, it is surprising that the evolution of porous structures exposed to erosion and deposition has only been partially understood both theoretically and experimentally.

In our work, we study a generic model of network evolution, that includes previous examples as particular cases. We show that depending on the erosion dynamics, the flow statistics may homogenize, stay unaffected, or become unstable and develop channels. We quantify the phase transition and using a simple model we identify quantitative criteria to distinguish between these regimes and as a result correctly predicting the fate of
the network, which we corroborate numerically.

Our results signify the importance of local dynamics and the nature of the feedback mechanism on the long-time global behavior of pore-networks and provide a simple route to connect local erosion dynamics to global bulk behaviors of the pore-network relevant for many environmental, biological, and industrial applications. We, therefore, trust that the results will be of interest to the broad readership of PRL.  We thank you for your consideration and look forward to hearing from you.

@@

\textbf{Karen Alim}, Max Planck Institute for dynamics and self-organization, G\"ottingen, Germany, \texttt{k.alim@tum.de}

\textbf{Martin Blunt}, Imperial College London, \texttt{m.blunt@imperial.ac.uk}

\textbf{Harvey Scher}, Weizmann Institute of Science, \texttt{harvey.scher@weizmann.ac.il}

\textbf{Martin Bazant}, Massachusetts Institute of Technology, \texttt{bazant@mit.edu}

\textbf{Arshad Kudrolli}, Clark University, \texttt{akudrolli@clarku.edu}
